---
title: "Code Challenge 1"
output: html_document
---

# CODE CHALLENGE 1 TEMPLATE

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1: Summary Statistics

### Question 1: Mean and Standard Deviation Calculation
#### Part 1: Calculate the mean of the following vector: `c(12, 15, 22, 29, 35)`.

```{r}
# Write your code here
numbers <- c(12, 15, 22, 29, 35)
mean_value <- ___
mean_value
```

#### Part 2: Calculate the standard deviation of the same vector.

```{r}
# Write your code here
std_dev_value <- ___
std_dev_value
```

## Task 2: Data Visualization with ggplot2

### Question 2: Scatter Plot and Linear Regression Line
#### Part 1: Create a scatter plot for the `mtcars` dataset with `wt` (weight) on the x-axis and `mpg` (miles per gallon) on the y-axis.

```{r}
# Load the ggplot2 library
library(ggplot2)

# Create scatter plot
plot <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  ggtitle("Scatter Plot of Weight vs. MPG")

# Display the plot
___
```

#### Part 2: Add a linear regression line to the scatter plot.

```{r}
# Add linear regression line
plot <- plot + geom_smooth(method = "lm", se = FALSE, color = "red")
___
```

## Task 3: Probability

### Question 3: Probability Calculation and Binomial Distribution
#### Part 1: Calculate the probability of getting a sum of 8 when rolling two six-sided dice.

```{r}
# Calculate probability
probability <- ___ / 36
probability
```

#### Part 2: Calculate the probability of getting at least 3 heads in 5 flips of a fair coin.

```{r}
# Calculate binomial probability
binom_prob <- pbinom(___, size = 5, prob = 0.5, lower.tail = FALSE)
binom_prob
```

## Task 4: Data Handling and Manipulation

### Question 4: Filtering and Summarizing Data
#### Part 1: Filter the `mtcars` dataset to only include cars with `hp` (horsepower) greater than 100 and summarize the average `mpg` (miles per gallon) of these cars.

```{r}
# Filter data
filtered_data <- mtcars[mtcars$hp > 100, ]
average_mpg <- mean(___)
average_mpg
```

#### Part 2: Calculate the sum of `hp` for the filtered dataset.

```{r}
# Sum of horsepower
total_hp <- sum(___)
total_hp
```

## Task 5: Custom Functions

### Question 5: Custom Function for Quartiles and Interquartile Range
#### Part 1: Write a custom function to calculate the quartiles (Q1, Q2, Q3) of a numeric vector.

```{r}
# Custom function for quartiles
calculate_quartiles <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q2 <- quantile(x, 0.50)
  Q3 <- quantile(x, 0.75)
  return(c(Q1 = Q1, Q2 = Q2, Q3 = Q3))
}

# Test the function
quartiles <- calculate_quartiles(c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100))
quartiles
```

#### Part 2: Write a custom function to calculate the interquartile range (IQR) using the quartiles.

```{r}
# Custom function for IQR
calculate_iqr <- function(x) {
  quartiles <- calculate_quartiles(x)
  IQR <- quartiles[3] - quartiles[1]
  return(IQR)
}

# Test the function
iqr_value <- calculate_iqr(c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100))
iqr_value
```


You are expected to fill in the blanks (represented by `___`) with their solutions. Each question is designed to have two parts, making a total of 10 points. This template covers more complex tasks such as summary statistics, data visualization with `ggplot2`, probability calculations, data handling and manipulation, and custom function creation.