# Data Structures and Data Handling in R (Light Focus)

## Overview of Vectors, Matrices, Lists, and Data Frames

Understanding and effectively manipulating basic data structures in R is foundational for all subsequent data manipulation and analysis tasks. This section will guide you through creating and working with vectors, matrices, lists, and data frames.

## Creating and Manipulating Vectors

Vectors are the simplest and most common data structure in R. They store an ordered collection of elements of the same type. Understanding vectors is crucial as they form the building blocks for more complex data structures.

```{r}
# Creating a numeric vector
numbers <- c(1, 2, 3, 4, 5)

# Adding an element to the vector
numbers <- c(numbers, 6)

# Accessing specific elements in a vector using their index
second_element <- numbers[2]  # Accesses the second element in the vector

# Creating a vector for temperature conversion
temperature_c <- c(23, 21, 24, 22, 20)
temperature_f <- temperature_c * 9 / 5 + 32  # Convert to Fahrenheit
print(temperature_f)
```

## Operations with Matrices and Arrays

Matrices are vectors with a dimension attribute. They are useful for handling two-dimensional data. Arrays extend this concept to multiple dimensions.

```{r}
# Creating a matrix using the matrix function
matrix_data <- matrix(1:9, nrow=3, ncol=3)

# Accessing a specific element (row 1, column 2)
element <- matrix_data[1, 2]

# Transposing a matrix (switch rows and columns)
transposed <- t(matrix_data)

# Multiplying first row elements by 2 to demonstrate manipulation
matrix_data[1, ] <- matrix_data[1, ] * 2
print(matrix_data)
print(transposed)
```

## Understanding Lists and Their Uses

Lists in R are a generic vector capable of holding elements of different types. This makes them extremely versatile for datasets that require a mix of numeric, character, and logical types.

```{r}
# Creating a list with mixed types
my_list <- list(name="John", scores=c(95, 82, 90), passed=TRUE)

# Accessing elements within a list using the $ operator
scores <- my_list$scores
print(scores)

# Adding a new element to the list to show its flexibility
my_list$address <- "123 Elm St"
print(my_list)
```

## Data Frames for Tabular Data

Data frames are used extensively in R for storing tabular data. They resemble a table in a relational database or an Excel spreadsheet, with each column potentially holding a different type of data.

```{r}
# Creating a data frame
students <- data.frame(
  name = c("Alice", "Bob", "Carol"),
  score = c(88, 93, 85),
  pass = c(TRUE, TRUE, FALSE)
)

# Accessing a specific column using the $ operator
names <- students$name

# Adding a new column to indicate if the student scored an A (score > 90)
students$grade <- ifelse(students$score > 90, "A", "B")
print(students)
```

## Data Handling Techniques with dplyr

The `dplyr` package in R is a powerful tool for data manipulation, providing a coherent set of verbs that help you solve the most common data manipulation challenges. This section will cover how to use these verbs to select, filter, arrange, mutate, summarize, join, and group your data.

## Selecting, Filtering, and Arranging Data

`dplyr` makes it easy to select specific columns of data, filter rows based on conditions, and arrange data in ascending or descending order.

### Selecting Data

Use `select()` to choose a subset of columns to keep.

```{r}
library(dplyr)

# Sample data frame
data <- data.frame(
  name = c("Alice", "Bob", "Carol", "David"),
  age = c(25, 30, 19, 22),
  salary = c(50000, 54000, 32000, 45000)
)

# Selecting only the name and age columns
selected_data <- select(data, name, age)
print(selected_data)
```

### Filtering Data

`filter()` is used to retrieve rows that meet certain conditions.

```{r}
# Filtering to find only those over 25 years old
filtered_data <- filter(data, age > 25)
print(filtered_data)
```

### Arranging Data

`arrange()` is used to reorder rows of a data frame.

```{r}
# Arranging data by age in ascending order
arranged_data <- arrange(data, age)
print(arranged_data)
```

## Mutating and Summarizing Data Sets

`mutate()` adds new variables that are functions of existing variables, and `summarise()` reduces multiple values down to a single summary.

### Mutating Data

```{r}
# Adding a new column that calculates yearly savings potential
mutated_data <- mutate(data, savings = salary * 0.1)
print(mutated_data)
```

### Summarizing Data

```{r}
# Summarizing to find the average salary
summary_data <- summarise(data, average_salary = mean(salary))
print(summary_data)
```

## Joins and Data Merging Techniques

`dplyr` provides several functions to merge data frames together, such as `left_join()`, `right_join()`, `inner_join()`, and `full_join()`.

```{r}
# Additional data frame with department information
department_data <- data.frame(
  name = c("Alice", "Bob", "David"),
  department = c("Finance", "IT", "Marketing")
)

# Joining data on the name column
joined_data <- left_join(data, department_data, by = "name")
print(joined_data)
```

## Group Operations with `group_by` and `summarise`

Grouping data makes it possible to compute summaries over groups of data.

```{r}
# Grouping data by department and calculating average salary
grouped_data <- data %>%
  left_join(department_data, by = "name") %>%
  group_by(department) %>%
  summarise(average_salary = mean(salary, na.rm = TRUE))

print(grouped_data)
```

This demonstration of `dplyr`'s capabilities shows how straightforward it is to manipulate and analyze data sets in R. These functions are not only powerful but also help make your data manipulation tasks more readable and concise.