# Data Visualization in R (Light Focus)

## Introduction to ggplot2 and Basic Plotting

Effective data visualization is key to interpreting data and communicating results clearly. In R, `ggplot2` is one of the most powerful tools for creating a wide variety of static, aesthetic, and informative plots. This section introduces `ggplot2`, its syntax, and basic plotting techniques.

### ggplot2 Syntax and Layering System

`ggplot2` is based on the grammar of graphics, a system that allows you to create graphs layer by layer. You start with data, add `aes` (aesthetics) to indicate which variables to plot, and then add `geoms` (geometric objects) to decide the type of plot.

```{r}
library(ggplot2)

# Basic syntax
ggplot(data = mtcars, aes(x = wt, y = mpg)) + 
  geom_point()  # adds a layer for scatter plot
```

In this example, `mtcars` is the dataset, `wt` (car weight) and `mpg` (miles per gallon) are the variables, and `geom_point()` creates a scatter plot.

### Creating Histograms, Bar Plots, and Scatter Plots

Histograms, bar plots, and scatter plots are fundamental for exploring data distributions and relationships between variables.

#### Histograms

Histograms are used to visualize the distribution of a single continuous variable by dividing the data into bins and counting the number of observations in each bin.

```{r}
# Creating a histogram of car weights
ggplot(mtcars, aes(x = wt)) + 
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black") +
  labs(title = "Histogram of Car Weights", x = "Weight (1000 lbs)", y = "Frequency")
```

#### Bar Plots

Bar plots are useful for comparing quantities corresponding to different groups.

```{r}
# Convert cyl (number of cylinders) to a factor for a bar plot
mtcars$cyl <- as.factor(mtcars$cyl)

# Creating a bar plot of car counts per cylinder type
ggplot(mtcars, aes(x = cyl)) +
  geom_bar(fill = "tomato", color = "black") +
  labs(title = "Number of Cars by Cylinder Type", x = "Cylinders", y = "Count")
```

#### Scatter Plots

Scatter plots are ideal for examining the relationship between two continuous variables.

```{r}
# Creating a scatter plot of mpg vs wt
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, color = "dodgerblue") +
  labs(title = "Car Weight vs. MPG", x = "Weight (1000 lbs)", y = "Miles per Gallon")
```

### Aesthetics and Themes

`ggplot2` allows extensive customization of plots through aesthetics and themes, enabling you to make plots more informative and appealing.

```{r}
# Enhancing scatter plot with themes and labels
ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +  # Color points by cylinder count
  geom_point(size = 4) +
  labs(title = "Car Weight vs. MPG",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon",
       color = "Number of Cylinders") +
  theme_minimal()  # Using a minimal theme
```

These examples demonstrate basic `ggplot2` usage for creating various types of plots, setting the stage for more advanced visualization techniques. Each type of plot serves different purposes and can be tailored extensively using `ggplot2`'s powerful customization options.

## Customizing Graphs for Data Presentation

Visualizing data effectively involves more than just creating basic plots; it requires customization to enhance clarity and impact. This section of the course will teach you how to customize axes, legends, and labels, fine-tune scales and coordinates, and add annotations and custom geometries using ggplot2 in R.

## Customizing Axes, Legends, and Labels

Customizing the textual components of your plots helps in making them more readable and informative. `ggplot2` allows for detailed adjustments to these elements.

### Customizing Axes

You can modify the text of axis labels and adjust the axis limits to focus on specific areas of the data.

```{r}
# Example of customizing axes
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  scale_x_continuous("Weight (in 1000 lbs)", limits = c(1, 6)) +
  scale_y_continuous("Miles Per Gallon", limits = c(10, 35)) +
  labs(title = "Car Weight vs. Fuel Efficiency")
```

### Customizing Legends

Adjusting the legend involves changing its title, text, and position to improve the overall aesthetics and readability.

```{r}
# Example of customizing legends
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() +
  labs(color = "Number of Cylinders") +
  theme(legend.position = "bottom")
```

### Customizing Labels

Labels for the entire plot, axes, and legends can be styled to enhance clarity or to align with publication standards or personal preferences.

```{r}
# Example of customizing labels
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() +
  labs(
    title = "Vehicle Dynamics",
    x = "Car Weight (1000 lbs)",
    y = "Fuel Efficiency (MPG)",
    color = "Cylinders"
  )
```

## Fine-tuning Scales and Coordinates

Manipulating scales and coordinates can help in presenting data more effectively by adjusting the visual representation to better fit the dataset's story.

### Scales

Scales adjust how data is mapped to aesthetics. You can set the scale for each aesthetic (like color, size, or shape) to control the visual properties of the plot.

```{r}
# Example of adjusting color scales
ggplot(mtcars, aes(x = wt, y = mpg, color = hp)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red")
```

### Coordinates

Using coordinate functions, you can adjust the plotting area, aspect ratios, or flip axes which is useful for certain types of data or specific visual effects.

```{r}
# Example of flipping coordinates
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  coord_flip()
```

## Adding Annotations and Custom Geometries

Annotations and custom geometries enhance plots by adding text annotations, lines, or other shapes to highlight specific features or data points.

### Annotations

Adding text annotations to highlight specific points or areas on the plot.

```{r}
# Example of adding annotations
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  annotate("text", x = 5, y = 30, label = "Outlier", color = "red")
```

### Custom Geometries

You can draw shapes or lines to highlight trends or important areas.

```{r}
# Example of adding custom geometries
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_hline(yintercept = 20, linetype = "dashed", color = "gray")
```

These customization techniques in ggplot2 enable you to tailor your visualizations precisely to your data’s narrative, improving not only the aesthetics but also the effectiveness of your data presentations.

## Interactive Demonstrations

Let's illustrate these concepts with interactive examples. Try running these scripts in your R environment to see how `ggplot2` brings your data to life.

### Interactive Example: Creating a Scatter Plot

```{r}
library(ggplot2)

# Creating a scatter plot of mpg vs wt
scatter_plot <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(size = 3, color = "dodgerblue") +
  labs(title = "Car Weight vs. MPG", x = "Weight (1000 lbs)", y = "Miles per Gallon")
print(scatter_plot)
```

### Interactive Example: Customizing a Scatter Plot

```{r}
# Enhancing scatter plot with themes and labels
enhanced_scatter_plot <- ggplot(mtcars, aes(x = wt, y = mpg, color = cyl)) +  # Color points by cylinder count
  geom_point(size = 4) +
  labs(title = "Car Weight vs. MPG",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon",
       color = "Number of Cylinders") +
  theme_minimal()  # Using a minimal theme
print(enhanced_scatter_plot)
```

### Interactive Example: Adding Annotations and Custom Geometries

```{r}
# Adding annotations and custom geometries to a scatter plot
annotated_plot <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_hline(yintercept = 20, linetype = "dashed", color = "gray") +
  annotate("text", x = 5, y = 30, label = "Outlier", color = "red") +
  labs(title = "Annotated Car Weight vs. MPG",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon")
print(annotated_plot)
```

By running these interactive examples, you will gain a deeper understanding of how to leverage `ggplot2` for creating effective and visually appealing data visualizations in R.