# Working with Different Types of Data (Light Focus)

## Overview of Qualitative and Quantitative Data

Understanding the distinction between qualitative and quantitative data is fundamental in data analysis, as it informs the appropriate methods for processing and analyzing data. This section provides a detailed explanation and demonstrations of these two major types of data.

### Defining Qualitative vs Quantitative Data

**Qualitative Data**  
Qualitative data, also known as categorical data, represents attributes, labels, or non-numeric entries. It describes qualities or characteristics that are observed and recorded. This type of data is often textual but may also include images or objects.

- **Examples**: Colors (red, blue, green), preferences (like, dislike), product types (household, industrial), or any labels categorizing data such as names of places or people.

**Demonstration of Qualitative Data**:
```{r}
# Sample qualitative data: Survey responses
survey_data <- data.frame(
  respondent_id = 1:3,
  favorite_color = c("Blue", "Red", "Green"),
  satisfaction_level = c("High", "Medium", "Low")
)

print(survey_data)
```

**Quantitative Data**  
Quantitative data involves numerical measurements or counts that can be quantified. This data type can be further classified as either discrete or continuous.

- **Discrete data**: Numeric data that has a countable number of values between any two values. A discrete variable cannot take the value of a fraction between one value and the next closest value.
  - **Example**: The number of users subscribed to a service, the number of cars in a parking lot.

- **Continuous data**: Numeric data that can take any value within a finite or infinite interval. It can represent measurements and therefore can take any value in an interval.
  - **Example**: Temperature, weight, height, and time.

**Demonstration of Quantitative Data**:
```{r}
# Sample quantitative data: Company sales data
sales_data <- data.frame(
  month = c("January", "February", "March"),
  revenue = c(10000, 15000, 13000),  # Continuous data
  units_sold = c(10, 15, 13)         # Discrete data
)

print(sales_data)
```

### Sampling Strategies and Data Collection Methods

When working with either qualitative or quantitative data, choosing the right sampling strategy and data collection method is crucial for obtaining reliable and representative data.

- **Random Sampling**: Ensures every member of the population has an equal chance of being selected. Reduces sampling bias.
- **Stratified Sampling**: Divides the population into smaller groups, or strata, based on shared characteristics before sampling. This is useful when the population is heterogeneous.

**Demonstration of Sampling Strategies**:
```{r}
# Assuming a larger dataset with multiple categories
data <- data.frame(
  category = rep(c("A", "B", "C"), times = 100),
  value = rnorm(300)
)

# Stratified sampling: Sample 10 from each category
sampled_data <- do.call(rbind, lapply(split(data, data$category), function(subdata) {
  subdata[sample(nrow(subdata), 10), ]
}))

print(sampled_data)
```

### Structuring Data Sets for Analysis

Properly structuring data sets is essential for analysis, particularly when dealing with mixed data types. Structured data ensures consistency and ease of access during analysis.

- **Data Frames**: R's primary data structure for storing datasets with multiple types of information.
- **Matrix**: Suitable for numerical data where operations like matrix multiplication are needed.

**Demonstration of Data Structuring**:
```{r}
# Creating a structured data frame
structured_data <- data.frame(
  ID = 1:5,
  Name = c("Alice", "Bob", "Carol", "David", "Eve"),
  Age = c(25, 30, 35, 40, 45),
  Salary = c(50000, 60000, 55000, 58000, 62000)
)

print(structured_data)
```

This section has introduced the foundational concepts of qualitative and quantitative data, illustrated with R examples. Understanding these distinctions and how to work with different data types prepares you for more effective data analysis and interpretation.

## Best Practices in Data Collection and Cleaning

Effective data collection and cleaning are crucial for ensuring the accuracy and reliability of data analysis. This section covers essential practices for identifying and handling missing data, performing data type conversions and formatting, and detecting and dealing with outliers.

## Identifying and Handling Missing Data

Missing data can significantly impact the results of your analysis. Properly identifying and handling missing entries is a critical step in data cleaning.

### Identifying Missing Data

You can identify missing data in R using functions like `is.na()` which returns a logical vector indicating which elements are NA.

```{r}
# Example dataset with missing values
data <- data.frame(
  names = c("Alice", "Bob", NA, "David"),
  scores = c(92, NA, 88, 94)
)

# Identifying missing data
missing_data <- is.na(data)
print(missing_data)
```

### Handling Missing Data

Once identified, you can handle missing data by either removing it or imputing it based on the context of your analysis.

- **Removing missing data**: Use `na.omit()` to remove rows with NA values.
- **Imputing missing data**: Replace missing values with statistical measures (mean, median) or more sophisticated algorithms.

```{r}
# Removing missing data
clean_data <- na.omit(data)
print(clean_data)

# Imputing missing data with the mean
data$scores[is.na(data$scores)] <- mean(data$scores, na.rm = TRUE)
print(data)
```

## Data Type Conversions and Formatting

Data often comes in formats that are not suitable for analysis. Converting data into the correct type is essential for further processing.

### Data Type Conversion

Use functions like `as.numeric()`, `as.factor()`, etc., to convert data types in R.

```{r}
# Example of converting character data to numeric
data$ages <- c("20", "25", "30", "35")
data$ages <- as.numeric(data$ages)
print(data$ages)
```

### Formatting Data

Formatting data for readability and consistency (e.g., date formats) ensures that your datasets are easy to understand and analyze.

```{r}
# Formatting dates
dates <- c("01-01-2020", "02-01-2020", "03-01-2020")
formatted_dates <- as.Date(dates, format = "%d-%m-%Y")
print(formatted_dates)
```

## Detecting and Dealing with Outliers

Outliers can skew the results of your data analysis, making it important to detect and appropriately handle them.

### Detecting Outliers

A common method to detect outliers is to use statistical thresholds like the interquartile range (IQR).

```{r}
# Detecting outliers using IQR
scores <- c(100, 102, 99, 105, 110, 200, 98, 97, 95, 250)
Q1 <- quantile(scores, 0.25)
Q3 <- quantile(scores, 0.75)
IQR <- Q3 - Q1
outliers <- scores < (Q1 - 1.5 * IQR) | scores > (Q3 + 1.5 * IQR)
print(scores[outliers])
```

### Handling Outliers

Handling outliers depends on their cause and the amount of impact they have on your analysis.

- **Removing outliers**: If they are errors or extremely rare.
- **Capping values**: Replacing outliers with the highest non-outlier values.

```{r}
# Removing outliers
clean_scores <- scores[!outliers]
print(clean_scores)

# Capping outliers
scores[scores < (Q1 - 1.5 * IQR)] <- Q1 - 1.5 * IQR
scores[scores > (Q3 + 1.5 * IQR)] <- Q3 + 1.5 * IQR
print(scores)
```

This section has covered essential practices for handling different types of data, from identifying and managing missing data to converting and formatting data types, and detecting and dealing with outliers. These practices are foundational for ensuring the integrity and reliability of your data analysis.

### Interactive Demonstrations

Let's illustrate these concepts with interactive examples. Try running these scripts in your R environment to see how they apply to real data.

### Interactive Example: Identifying and Handling Missing Data

```{r}
# Provided data frame
data <- data.frame(
  name = c("Alice", "Bob", NA, "David"),
  score = c(85, NA, 88, 90)
)

# Identify missing values
print("Missing Data Identification:")
print(is.na(data))

# Fill missing scores with the column mean
data$score[is.na(data$score)] <- mean(data$score, na.rm = TRUE)
print("Data after handling missing values:")
print(data)
```

### Interactive Example: Converting Data Types

```{r}
# Character data
char_data <- data.frame(
  values = c("100", "200", "300", "400")
)

# Convert character to numeric
char_data$values <- as.numeric(char_data$values)

# Print the converted data and data type
print("Converted Data:")
print(char_data)
print("Data Type Verification:")
print(sapply(char_data, class))
```

### Interactive Example: Detecting and Handling Outliers

```{r}
# Provided data
values <- c(50, 55, 45, 60, 400, 65, 50, 430, 49, 52)

# Detect outliers
Q1 <- quantile(values, 0.25)
Q3 <- quantile(values, 0.75)
IQR <- Q3 - Q1
outliers <- which(values < (Q1 - 1.5 * IQR) | values > (Q3 + 1.5 * IQR))

# Replace outliers with the median
values[outliers] <- median(values)

# Print the cleaned data
print("Cleaned Data:")
print(values)
```

### Interactive Example: Date Formatting

```{r}
# Provided date strings
date_strings <- c("01-12-2022", "15-11-2022", "23-10-2022")

# Convert to Date object
dates <- as.Date(date_strings, format = "%d-%m-%Y")

# Reformat dates to "yyyy/mm/dd"
formatted_dates <- format(dates, "%Y/%m/%d")

# Print the formatted dates
print("Formatted Dates:")
print(formatted_dates)
```

These interactive examples demonstrate how to handle common data preprocessing tasks, ensuring your data is clean, correctly formatted, and ready for analysis.