# Categorical Data Analysis

## Introduction to Logistic Regression

Logistic regression is used when the dependent variable is binary (i.e., it has two possible outcomes). It models the probability of the occurrence of an event by fitting data to a logistic curve.

### Odds Ratios and Logit Function
- **Odds Ratios**: The odds ratio is a measure of association between an exposure and an outcome. It represents the odds that an event will occur given a particular exposure, compared to the odds of the event occurring without that exposure.
- **Logit Function**: The logit function is the natural logarithm of the odds of the dependent event. Logistic regression uses this function to model the relationship between the independent variables and the probability of the dependent event occurring.

### Model Fitting and Interpretation
- **Model Fitting**: Logistic regression models are fitted using maximum likelihood estimation. The coefficients obtained from the model represent the change in the log odds of the dependent variable for a one-unit change in the independent variable.
- **Interpretation**: The exponentiated coefficients (exp(coef)) can be interpreted as odds ratios. A coefficient of 0 means no effect, positive coefficients increase the odds, and negative coefficients decrease the odds.

**Example and Demonstration**:

```{r}
# Example data
set.seed(123)
x1 <- rnorm(100)
x2 <- rnorm(100)
y <- rbinom(100, 1, prob = 1 / (1 + exp(-(0.5 + 1.5 * x1 - 1 * x2))))

data <- data.frame(y, x1, x2)

# Fit Logistic Regression
model_logit <- glm(y ~ x1 + x2, data = data, family = binomial)
summary(model_logit)

# Interpretation of coefficients
exp(coef(model_logit))
```

### Assessing Model Goodness of Fit
- **Goodness of Fit**: The goodness of fit of a logistic regression model can be assessed using various methods such as the Hosmer-Lemeshow test, Akaike Information Criterion (AIC), and analysis of residuals.
- **Example**: Use the Hosmer-Lemeshow test to assess the model fit.

```{r}
# Assessing model goodness of fit
library(ResourceSelection)
hoslem.test(data$y, fitted(model_logit))
```

## Modeling and Interpretation of Binary Outcomes

### Predictive Modeling with Binary Data
Logistic regression is commonly used for predictive modeling with binary outcomes. The model predicts the probability that a given input belongs to one of the two outcome categories.

### Evaluating Model Performance (ROC Curves, AUC)
- **ROC Curve**: A Receiver Operating Characteristic (ROC) curve is a graphical representation of a model's diagnostic ability. It plots the true positive rate (sensitivity) against the false positive rate (1 - specificity) for different threshold values.
- **AUC (Area Under the Curve)**: The AUC is a single scalar value that summarizes the performance of the model across all threshold values. A higher AUC indicates better model performance.

**Example and Demonstration**:

```{r}
# ROC Curve and AUC
library(pROC)
roc_curve <- roc(data$y, fitted(model_logit))
plot(roc_curve)
auc(roc_curve)
```

## Comprehensive Exercise: Logistic Regression with the `iris` Dataset

### Exercise: Logistic Regression Analysis

**Objective**: Use logistic regression to analyze the binary outcome variable in the `iris` dataset. The exercise will include fitting the model, interpreting the results, assessing goodness of fit, and evaluating model performance.

**Steps**:
1. **Fit a logistic regression model**.
2. **Interpret the coefficients**.
3. **Assess model goodness of fit**.
4. **Evaluate model performance using ROC curves and AUC**.

### Detailed Steps

```{r}
# Load necessary libraries
library(ggplot2)
library(caret)
library(dplyr)
library(pROC)
library(ResourceSelection)

# Load the dataset and create a binary outcome variable
data <- iris
data$Species <- ifelse(data$Species == "setosa", 1, 0)

# 1. Fit Logistic Regression
model_logit <- glm(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, 
                   data = data, family = binomial)
summary(model_logit)

# 2. Interpretation of coefficients
exp(coef(model_logit))

# 3. Assessing model goodness of fit
hoslem_test <- hoslem.test(data$Species, fitted(model_logit))
print(hoslem_test)

# 4. ROC Curve and AUC
roc_curve <- roc(data$Species, fitted(model_logit))
plot(roc_curve)
auc(roc_curve)

# Comprehensive script
logistic_regression_analysis <- function(data) {
  # Fit Logistic Regression
  model_logit <- glm(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, 
                     data = data, family = binomial)
  print(summary(model_logit))
  
  # Interpretation of coefficients
  print(exp(coef(model_logit)))
  
  # Assessing model goodness of fit
  hoslem_test <- hoslem.test(data$Species, fitted(model_logit))
  print(hoslem_test)
  
  # ROC Curve and AUC
  roc_curve <- roc(data$Species, fitted(model_logit))
  plot(roc_curve)
  print(auc(roc_curve))
}

# Run the comprehensive analysis
logistic_regression_analysis(data)
```