# Advanced Data Manipulation and Transformation (`OPTIONAL`)

In advanced data analysis, efficiently manipulating and transforming data is crucial. The `tidyverse` package in R provides a powerful and coherent set of tools for data wrangling.

By following this comprehensive exercise, students will gain hands-on experience with advanced data manipulation and transformation techniques using the `tidyverse` package in R. This will prepare them for tackling complex data wrangling tasks in their data analysis projects.

## Tidyverse Approaches to Data Wrangling

The `tidyverse` is a collection of R packages designed for data science. It includes `dplyr` for data manipulation, `tidyr` for data tidying, and `ggplot2` for data visualization.

### Key Functions in `dplyr`
- **`select()`**: Select specific columns.
- **`filter()`**: Filter rows based on conditions.
- **`mutate()`**: Create new columns or modify existing ones.
- **`summarise()`**: Summarize data (e.g., calculating means or counts).
- **`group_by()`**: Group data by one or more variables.

**Example**:

```{r}
library(tidyverse)

# Load the dataset
data <- iris

# Select specific columns
selected_data <- data %>%
  select(Sepal.Length, Species)

# Filter rows based on condition
filtered_data <- data %>%
  filter(Sepal.Length > 5)

# Create a new column
mutated_data <- data %>%
  mutate(Sepal.Ratio = Sepal.Length / Sepal.Width)

# Summarize data
summary_data <- data %>%
  group_by(Species) %>%
  summarise(Mean_Sepal.Length = mean(Sepal.Length),
            Count = n())
```

## Dealing with Text and Categorical Data

Handling text and categorical data involves operations like re-coding factors, string manipulation, and one-hot encoding.

### Key Functions
- **`factor()`**: Create or modify factors.
- **`str_*` functions**: String manipulation functions from the `stringr` package (part of `tidyverse`).

**Example**:

```{r}
# Create a factor variable
data$Species <- factor(data$Species, levels = c("setosa", "versicolor", "virginica"))

# String manipulation
library(stringr)
text_data <- c("apple", "banana", "cherry")
text_lengths <- str_length(text_data)
text_upper <- str_to_upper(text_data)
```

## Working with Dates and Times

Handling dates and times is critical in many data analyses. The `lubridate` package (part of the `tidyverse`) makes it easier to work with date-time data.

### Key Functions
- **`ymd()`, `mdy()`, `dmy()`**: Parse dates in different formats.
- **`year()`, `month()`, `day()`**: Extract components of dates.
- **`now()`**: Get the current date and time.

**Example**:

```{r}
library(lubridate)

# Parse dates
dates <- c("2023-01-01", "2023-02-15", "2023-03-10")
parsed_dates <- ymd(dates)

# Extract components
years <- year(parsed_dates)
months <- month(parsed_dates)
days <- day(parsed_dates)

# Get current date and time
current_time <- now()
```

# Comprehensive Exercise: Data Wrangling and Transformation with `tidyverse`

## Exercise: Advanced Data Manipulation with the `mtcars` Dataset

**Objective**: Use `tidyverse` functions to perform advanced data manipulation and transformation tasks on the `mtcars` dataset.

**Steps**:
1. **Select and rename columns**.
2. **Filter rows based on multiple conditions**.
3. **Create new columns and modify existing ones**.
4. **Group data and calculate summary statistics**.
5. **Handle categorical data and perform string manipulations**.
6. **Work with date-time data**.

## Detailed Steps

```{r}
# Load necessary libraries
library(tidyverse)
library(lubridate)

# Load the dataset
data <- mtcars

# 1. Select and rename columns
selected_data <- data %>%
  select(mpg, cyl, hp, wt) %>%
  rename(Miles_Per_Gallon = mpg, 
         Cylinders = cyl, 
         Horsepower = hp, 
         Weight = wt)

# 2. Filter rows based on multiple conditions
filtered_data <- selected_data %>%
  filter(Miles_Per_Gallon > 20, Cylinders == 4)

# 3. Create new columns and modify existing ones
mutated_data <- filtered_data %>%
  mutate(Horsepower_per_Weight = Horsepower / Weight,
         Weight_Category = ifelse(Weight > 3, "Heavy", "Light"))

# 4. Group data and calculate summary statistics
summary_data <- mutated_data %>%
  group_by(Weight_Category) %>%
  summarise(Mean_Horsepower_per_Weight = mean(Horsepower_per_Weight),
            Count = n())

# 5. Handle categorical data and perform string manipulations
# Create a factor variable
mutated_data$Weight_Category <- factor(mutated_data$Weight_Category)

# String manipulation example
car_names <- rownames(data)
car_name_lengths <- str_length(car_names)
car_name_upper <- str_to_upper(car_names)

# 6. Work with date-time data
# Create a sample date column
mutated_data <- mutated_data %>%
  mutate(Sample_Date = ymd("2023-05-01") + days(1:n()))

# Extract components of dates
mutated_data <- mutated_data %>%
  mutate(Year = year(Sample_Date),
         Month = month(Sample_Date),
         Day = day(Sample_Date))

# Comprehensive script
mtcars_analysis <- function() {
  # Load and manipulate data
  data <- mtcars
  
  # Select and rename columns
  selected_data <- data %>%
    select(mpg, cyl, hp, wt) %>%
    rename(Miles_Per_Gallon = mpg, 
           Cylinders = cyl, 
           Horsepower = hp, 
           Weight = wt)

  # Filter rows based on conditions
  filtered_data <- selected_data %>%
    filter(Miles_Per_Gallon > 20, Cylinders == 4)

  # Create and modify columns
  mutated_data <- filtered_data %>%
    mutate(Horsepower_per_Weight = Horsepower / Weight,
           Weight_Category = ifelse(Weight > 3, "Heavy", "Light"))

  # Group and summarize data
  summary_data <- mutated_data %>%
    group_by(Weight_Category) %>%
    summarise(Mean_Horsepower_per_Weight = mean(Horsepower_per_Weight),
              Count = n())

  print("Summary Data:")
  print(summary_data)

  # String manipulation
  car_names <- rownames(data)
  car_name_lengths <- str_length(car_names)
  car_name_upper <- str_to_upper(car_names)
  print("Car Names in Upper Case:")
  print(car_name_upper)

  # Work with date-time data
  mutated_data <- mutated_data %>%
    mutate(Sample_Date = ymd("2023-05-01") + days(1:n()),
           Year = year(Sample_Date),
           Month = month(Sample_Date),
           Day = day(Sample_Date))
  
  print("Mutated Data with Dates:")
  print(mutated_data)
}

# Run the comprehensive analysis
mtcars_analysis()
```
