# Advanced R Programming Techniques (`OPTIONAL`)

## Writing Functions and Loops for Automating Tasks

In advanced R programming, writing functions and loops efficiently is crucial for automating repetitive tasks and handling complex data operations.

### Function Arguments and Return Values
- **Function Arguments**: Functions can take multiple arguments, including default values, to make them more flexible and reusable.
- **Return Values**: Functions can return a single value or multiple values as a list.

**Example**:

```{r}
# Define a function with arguments and return value
my_function <- function(x, y = 2) {
  result <- x * y
  return(result)
}

# Call the function
my_function(5)    # Uses default value for y
my_function(5, 3) # Uses provided value for y
```

### Looping Constructs: `for`, `while`, and `apply` Family
- **`for` Loop**: Iterates over a sequence of elements.
- **`while` Loop**: Continues to execute as long as a condition is true.
- **`apply` Family**: Functions like `lapply`, `sapply`, `tapply`, and `mapply` apply a function over elements of a list or vector, offering more concise and often faster alternatives to loops.

**Example**:

```{r}
# for loop example
for (i in 1:5) {
  print(i)
}

# while loop example
i <- 1
while (i <= 5) {
  print(i)
  i <- i + 1
}

# apply family example
vec <- 1:5
squared <- sapply(vec, function(x) x^2)
print(squared)
```

### Avoiding Loops: Vectorization and Parallel Processing
- **Vectorization**: Writing code that operates on entire vectors or matrices at once, which is usually faster and more efficient than using loops.
- **Parallel Processing**: Using multiple cores or processors to perform computations simultaneously, which can significantly speed up data processing tasks.

**Example**:

```{r}
# Vectorized operation
vec <- 1:5
squared_vec <- vec^2
print(squared_vec)

# Parallel processing example using `parallel` package
library(parallel)
cl <- makeCluster(detectCores() - 1)
parSapply(cl, 1:5, function(x) x^2)
stopCluster(cl)
```

## Comprehensive Exercise: Automating Data Analysis with Functions and Loops

### Exercise: Automating Data Analysis with the `iris` Dataset

**Objective**: Write functions and use loops to automate the process of summarizing and analyzing the `iris` dataset. The exercise will include calculating summary statistics, visualizing data, and applying a machine learning model.

**Steps**:
1. **Write a function to calculate summary statistics for each species**.
2. **Use a loop to generate boxplots for each variable by species**.
3. **Write a function to train a decision tree model and evaluate its accuracy**.
4. **Combine these tasks into a comprehensive script that automates the entire process**.

### Detailed Steps

```{r}
# Load necessary libraries
library(ggplot2)
library(caret)
library(rpart)
library(dplyr)

# 1. Function to calculate summary statistics for each species
summary_stats <- function(data, species) {
  data %>% 
    filter(Species == species) %>%
    summarise(across(where(is.numeric), list(mean = mean, sd = sd)))
}

# 2. Loop to generate boxplots for each variable by species
plot_boxplots <- function(data) {
  variables <- colnames(data)[1:4]
  for (var in variables) {
    p <- ggplot(data, aes_string(x = "Species", y = var)) + 
      geom_boxplot() +
      ggtitle(paste("Boxplot of", var, "by Species"))
    print(p)
  }
}

# 3. Function to train a decision tree model and evaluate its accuracy
train_decision_tree <- function(data) {
  set.seed(123)
  train_index <- createDataPartition(data$Species, p = 0.8, list = FALSE)
  train_data <- data[train_index, ]
  test_data <- data[-train_index, ]
  
  model <- rpart(Species ~ ., data = train_data, method = "class")
  predictions <- predict(model, test_data, type = "class")
  accuracy <- confusionMatrix(predictions, test_data$Species)$overall['Accuracy']
  
  return(accuracy)
}

# 4. Comprehensive script
iris_analysis <- function() {
  data <- iris
  
  # Calculate summary statistics for each species
  species <- unique(data$Species)
  for (sp in species) {
    cat("\nSummary statistics for", sp, ":\n")
    print(summary_stats(data, sp))
  }
  
  # Generate boxplots
  plot_boxplots(data)
  
  # Train decision tree model and evaluate accuracy
  accuracy <- train_decision_tree(data)
  cat("\nDecision tree model accuracy:", accuracy, "\n")
}

# Run the comprehensive analysis
iris_analysis()
```